---
title: "03_Clustering"
author: "Rodrigo Gularte Mérida"
date: "9/6/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup
```{r libraries}
suppressPackageStartupMessages({
    library(Seurat)
    library(tidyverse)
    library(patchwork)
    devtools::load_all()
})
```

## load curated data
```{r data_load}
data(fibroS)

```

## Running Seurat Processing and clustering
* The SCTransform performs 3 crucial steps: Normalization, Finding Variable Genes and Scaling.

```{r seurat_pipeline}
set.seed(2022)
fibroS <- fibroS %>%
  SCTransform(vars.to.regress = "percent_mt", verbose = FALSE)
```

* Linear dimentionality reduction suggest between 16-20 PCs explain most of the data
```{r pca}
fibroS <- fibroS %>% RunPCA(npcs = 100)

ElbowPlot(fibroS, ndims = 50) + theme_light()
```

* *Note*: an alternate strategy is to estimate the proportion of variance explained.  In this case, 80% of the total variance is explaied by 34 PCs.
```{r variance_explained}
## calculate proportion of variance explained and choose threshold
prop_varexp <- fibroS@reductions$pca@stdev^2 / sum(fibroS@reductions$pca@stdev^2)

thrs <- c(80, 85, 90, 95)/100
pcs_at_thrs <- c(which(cumsum(prop_varexp) >= 0.80)[1],
                 which(cumsum(prop_varexp) >= 0.85)[1],
                 which(cumsum(prop_varexp) >= 0.90)[1],
                 which(cumsum(prop_varexp) >= 0.95)[1])

## variance explained plots
plot(cumsum(prop_varexp))
abline(h = thrs, col = "gray", lty = 3)
abline(v = pcs_at_thrs, col = "gray20", lty = 2)
text(x = pcs_at_thrs, y = rep(0.2, 4), labels = pcs_at_thrs, pos = 4)

```

```{r use_pc_and_clustering}
## use 80% variance explaned
pc_use <- pcs_at_thrs[1]

fibroS <- fibroS %>% 
  FindNeighbors(dims = 1:pc_use) %>%
  FindClusters(resolution = c(0.2, 0.4, 0.6, 0.8, 1)) %>%
  RunUMAP(dims = 1:pc_use)

```

```{r plot_check}
fibroS %>% DimPlot() + theme_light()

xx <- fibroS %>% 
  DimPlot(group.by = c("SCT_snn_res.0.2", "SCT_snn_res.0.4", "SCT_snn_res.0.6", "SCT_snn_res.0.8", "SCT_snn_res.1"), 
          ncol = 5)

for(i in 1:4) {
  xx[[i]] <- xx[[i]] + theme_light() + NoLegend() + theme(axis.title.x = element_blank(), axis.title.y = element_blank())
}
xx[[5]] <- xx[[5]] + theme_light()

```


```{r}
data(soleS)

soleS %>% DimPlot(group.by = "integrated_snn_res.0.4")

tmp1 <- merge(fibroS[[]], soleS[[]], by = 0, sort = FALSE) %>% select(SCT_snn_res.0.2, SCT_snn_res.0.4, SCT_snn_res.0.6, SCT_snn_res.0.8, SCT_snn_res.1, subj.y, age.y, integrated_snn_res.0.4, celltype.age )

tmp1 %>% select(SCT_snn_res.0.6, integrated_snn_res.0.4) %>% table

```
